Finding and controling biases in dark matter halo concentrations estimates

### 2016-07-05 paper/ms.tex
new structure

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.48\textwidth]{avg_foff_bolshoi.pdf}
  \includegraphics[width=0.48\textwidth]{avg_foff_vialactea.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Average value of the fractional offset ($f_{\rm
    off}$Eq. \ref{eq:f_off}) between the concentration in a halo at
  high resolution versus its downsampled versions.  Each point is the
  result of 100 downsampling iterations.  The velocity method
  noticeably overestimates the concentration up to a factor of $0.25$,
  while the new method only underestimates the concentrations by up to
  a factor of $0.05$.
    \label{fig:downsampling}}
\end{figure*}

@@

We divide the enclosed mass mass $M_i$ and the radii $r_i$ by their
virial values to obtain the dimensionless variables $m_i$ and $x_i$.
Once the mass profile is expressed in dimensionless variables the
concentration is the only free parameter.  We then use an Affine
Invariant Markov Chain Monte Carlo implemented in the python module
{\em emcee} \citep{emcee} to sample the likelihood function
distribution defined by ${\cal L}(c)\propto \exp(-\chi^2(c)/2)$ where
the $\chi^2(c)$ is written as

@@

We apply the three methods mentioned above on two different halo
samples.  The first sample is composed by mock halos generated to have
known concentration values in perfect spherical symmetry following an
NFW profile.  Using this sample we quantify the ability to recover the
expected concentration as a function of particle number.  The second
sample comes from a publicly available N-body cosmological simulation
to test the new method on more realistic and complex data to show the
impact on the mass-concentration relationship.

\subsection{Retrieving known concentrations from a mock halo sample}

\begin{figure}
\begin{center}
  \includegraphics[width=0.50\textwidth]{error.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Average value of the relative error in the concentration
  estimate, $\avg{|D|}$, as a function of the particle number in
  the set of mock halos. Different symbols represent different
  methods. The new method provides the most accurate estimate at fixed
  particle number.
    \label{fig:error}}
\end{figure}

The method we use to generate the mock halos is based on the
integrated mass profile.  We start by fixing the desired concentration
$c$ and total number of particles $N$ in the mock halo.  With these
values we define the mass element as $\delta m = 1/M$, corresponding
to the mass of each particle such that the total mass is one.  Then
for each particle, $i=1,\ldots,N$, we find the value of $r_i$ such
that the difference
% 
\begin{equation}
m(<r_i;c) - i \cdot \delta m
\end{equation}
%
is zero using Ridders' method.

The value of $r_i$ is the radius of the $i$-th particle of the mock
halo.  Then we generate random polar and azimuthal angles $\theta$ and
$\phi$ for each particle to ensure spherical symmetry.  Finally these
three spherical coordinates are transformed into Cartesian coordinates
$(r,\theta,\phi) \rightarrow (x,y,z)$.

We generate in total $400$ mock halos split into four different groups
of $100$ halos each.  The four groups differ in the total number of
particles for their halos: $20$, $200$, $2000$ and $20000$.  Inside
each group the halos have random concentration values in the range
$1<c<20$ with a uniform distribution.  We find the concentration
values for all these halos using the density, velocity and integrated
mass methods described in the previous section.

We quantify the difference between the expected $c_{in}$ and measured
$c_{out}$ values by
%
\begin{equation}
D= c_{out}/c_{in}-1,
\label{eq:D}
\end{equation}
%
and
%
\begin{equation}
\avg{|D|}=\frac{1}{\left|{\cal{H}}\right|}\sum_{{\cal{H}}} |D|,
\end{equation}
%
where ${\cal{H}}$ corresponds to a set of mock halos, and
$\left|{\cal{H}}\right|$ is the number of haloes in ${\cal{H}}$.  A
large discrepancy between the estimated and input concentrations are
quantified by a large $\avg{|D|}$.

Figure \ref{fig:error} shows the behaviour of $\avg{|D|}$ as a
function of halo particle number for the three different methods.

At fixed particle numbers the new method always shows the lowest
$\avg{|D|}$ values compared to the other two.  Its accuracy is on the
order of $10\%$ for $20$ particles in the halo, decreasing to $0.01\%$
for halos with $20000$ particles.  The dependence of $\avg{|D|}$ with
the particle number $N$ goes approximately as $\avg{|D|}\propto
N^{-1}$.

The method based on the maximum of the circular velocity shows a
similar behaviour with $\avg{|D|}\propto N^{-1/3}$.  Its accuracy is
$2-5$ times lower than in the new method, on the order of $20\%$ for
$20$ particle halos and $0.5\%$ for $20000$ particle halos.  The
method based on the direct density fit shows the lowest accuracy for a
low particle number.  As it is discussed in \citep{Munoz2011}, density
binning for halos with particle numbers below $\sim200$ leads to a
biased estimation of the mass density profile, and therefore to a
biased estimation of the concentration parameter.  This behaviour is
evident in the large values of $\avg{|D|}$ for halos with number of
particles below $\sim 200$ and an intermediate accuracy between the
other two methods for a high particle number.

We also find that the bias from different concentrations at fixed
particle number is negligible with respect to the effects induced by
different particle numbers.

\begin{figure}
\begin{center}
  \includegraphics[width=0.47\textwidth]{error_conc_bolshoi.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Uncertainties on different concentration estimates.  The
  large symbols show the standard deviation of the fractional offset
  ($f_{\rm off}$ as defined in Eq. \ref{eq:f_off}) from the downsampling
  experiments.  The small dots correspond to the $1\sigma$
  uncertainty on the concentration for each halo as estimated by the
  MCMC algorithm used for the new method.
  The solid lines shows the $1/N$ tendency, which also corresponds
  to the particle number dependence of the $\langle|D|\rangle$ statistic
  (i.e. the accuracy to retrieve a known concentration in mock halos)
  of the new method.
  Uncertainties due to particle number fluctuations are always larger
  than the statistical uncertainties from the fitting process. 
  \label{fig:downsampling_err}}
\end{figure}

\subsection{Measuring the concentration in halos from a cosmological simulation}
\label{subsec:cosmo}

The results presented in the previous section show that in an
idealized setup (pure NFW density profiles, perfect spherical halos in
total isolation) the new method performs better than the other two
methods commonly used in the literature.

However, simulated dark matter halos in a cosmological context are
complex.  They are not isolated, many of them experience mergers that
take them out of the equilibrium.  Halos may have also plenty of
substructure.  All these effects disturb the mass density profile,
taking it away from the ideal NFW profile.  In order to explore the
performance of the new method in this setting, we use it to estimate
the concentration parameter in dark matter halos obtained from a
cosmological N-body simulation and compare it with the other two
classical methods.

@@

\section{Bootstrapping to estimate biases}
\label{sec:bootstrapping}

@@

\subsubsection{Mass-Concentration Relationship}

>>>>

\begin{figure*}
\begin{center}
  \includegraphics[width=0.49\textwidth]{avg_foff_bolshoi.pdf}
  \includegraphics[width=0.48\textwidth]{sigma_foff_bolshoi.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Average value of the fractional offset ($f_{\rm
    off}$Eq. \ref{eq:f_off}) between the concentration in a halo at
  high resolution versus its downsampled versions.  Each point is the
  result of 100 downsampling iterations.  The velocity method
  noticeably overestimates the concentration up to a factor of $0.25$,
  while the new method only underestimates the concentrations by up to
  a factor of $0.05$.
Uncertainties on different concentration estimates.  The
  large symbols show the standard deviation of the fractional offset
  ($f_{\rm off}$ as defined in Eq. \ref{eq:f_off}) from the downsampling
  experiments.  The small dots correspond to the $1\sigma$
  uncertainty on the concentration for each halo as estimated by the
  MCMC algorithm used for the new method.
  The solid lines shows the $1/N$ tendency, which also corresponds
  to the particle number dependence of the $\langle|D|\rangle$ statistic
  (i.e. the accuracy to retrieve a known concentration in mock halos)
  of the new method.
  Uncertainties due to particle number fluctuations are always larger
  than the statistical uncertainties from the fitting process. 
    \label{fig:downsampling}}
\end{figure*}

@@

@@

We divide the enclosed mass mass $M_i$ and the radii $r_i$ by their
virial values to obtain the dimensionless variables $m_i$ and $x_i$.
Once the mass profile is expressed in dimensionless variables the
concentration is the only free parameter.  We then use an Affine
Invariant Markov Chain Monte Carlo implemented in the python module
{\texttt{emcee}} \citep{emcee} to sample the likelihood function
distribution defined by ${\cal L}(c)\propto \exp(-\chi^2(c)/2)$ where
the $\chi^2(c)$ is written as

@@

\section{Numerical Simulations and Halo Samples}

@@

\subsection{Bootstrapping to estimate biases}
\label{sec:bootstrapping}

@@

\subsection{Impact on the Mass-Concentration Relationship}
